### 2022-09-03 chapters/appendices/benchmarking_paper.tex

\section{Introduction}
%\lettrine[nindent=0em,lines=3]{C} 
To conduct robust evaluations, practitioners often try to ensure reproducible environmental conditions in order to properly benchmark their software systems.
In this area, reproducibility might be achieved by ensuring the same execution settings of physical nodes, virtual machines, clusters or cloud environments.
Recently, the research community has been investigating typical "crimes" in systems benchmarking and established guidelines for conducting robust and reproducible evaluations~\cite{DBLP:journals/corr/abs-1801-02381}.

In theory, using identical CPU, same memory configuration, similar storage and networking capabilities, should favour reproducible experiments.
However, when it comes to measuring the energy consumption of a system, applying acknowledged guidelines and carefully repeating the same benchmark can nonetheless lead to different energy footprints not only on homogeneous nodes, but even within a single node.
This difference---also called \emph{energy variation} (EV)---has become a serious threat to the accuracy of experimental evaluations.

Figure~\ref{fig:motivation} illustrates this variation problem as a violin plot of $20$ executions of the benchmark \emph{Conjugate Gradient} (\textsf{CG}) taken from the \emph{NAS Parallel Benchmarks} (NBP) suite~\cite{Bailey:1991:NPB:125826.125925}, on $4$ nodes of an homogeneous cluster (the cluster \textsf{Dahu} described in Table~\ref{table:g5k}) at 50\,\% workload.
One can observe a large variation of the energy consumption, not only among homogeneous nodes, but also at the scale of a single node, reaching up to $25\,\%$ in this example.

\begin{figure}%[!htb]
    \center{\includegraphics[width=.9\linewidth]{imgs/motivation}}
    \caption{CPU energy variation for the benchmark \textsf{CG}}\label{fig:motivation}
\end{figure}

@@

Unfortunately, hardware factors can be hardly tuned to tame the energy variation that can be observed in systems benchmarking.
For example, managing the CPU temperature or a server position in a cluster, are not actions that one can easily do, especially on the modern data centers and cloud platforms.
Therefore, the goal of this paper is to investigate the spectrum of factors that can cause or increase the variability of energy consumption in systems benchmarking, and to propose effective guidelines to control such factors in order to mitigate this variability.
While this paper does not question the benefits of established CPU features, like C-states or Turbo~Boost, it delivers a deeper analysis of the effects they may introduce on a wide set of experiments and nodes.
By quantifying potential energy variations induced by controllable factors, we intend to identify the proper configurations that minimize energy variations, depending on the workload characteristics.
These guidelines aim at supporting practitioners in conducting robust system benchmarks and reporting reproducible energy consumption, rather than recommending production-scale practices or solutions to reduce the power consumption of a given system.
% 
% The key contributions of this paper can therefore be summarized as:
% \begin{enumerate}
%     \item Providing a better understanding of the energy variation, by using different generations of CPU deployed in $4$ clusters with more than $100$ physical nodes, and by considering existing systems benchmarks with diverse workloads;
%     \item Identifying controllable factors that contribute to the variation in CPU energy consumption, comparing them against the state of the art, and completing them with other uncovered assumptions;
%     \item Reporting on some guidelines on how to conduct reproducible experiments with less energy variations;
%     \item Discussing the differences between inter-nodes and intra-nodes variations.
% \end{enumerate}

@@

Beyond hardware components, the accuracy of power meters has also been questioned.
Inadomi~\emph{et~al.}~\cite{inadomi_analyzing_2015} used three different power measurement tools: RAPL, Power Insight\footnote{\url{https://www.itssolution.com/products/trellis-power-insight-application}} and BGQ EMON.
All of the three tools recorded the same $10\,\%$ of energy variation, that was supposedly related to the manufacturing process.

\paragraph{Mitigating Energy Variations.}
Acknowledging the energy variation problem on processors, some papers proposed contributions to reduce and mitigate this variation.
In~\cite{inadomi_analyzing_2015}, the authors introduced a variation-aware algorithm that improves application performance under a power constraint by determining module-level (individual processor and associated DRAM) power allocation, with up to $5.4\times$ speedup.
The authors of~\cite{hammouda_noise-tolerant_2015} proposed parallel algorithms that tolerate the variability and the non-uniformity by decoupling per process communication over the available CPU.
Acun~\emph{et~al.}~\cite{acun_variation_2016} found out a way to reduce the energy variation on Ivy~Bridge and Sandy~Bridge processors, by disabling the Turbo~Boost feature to stabilize the execution time over a set of processors.
They also proposed some guidelines to reduce this variation by replacing the old slow chips, by load balancing the workload on the CPU cores and leaving one core idle.
They claimed that the variation between the processor cores is insignificant.
In~\cite{chasapis_runtime-guided_2016}, the researchers showed how a parallel system can be used to deal with the energy variation by compensating the uneven effects of power capping.

@@

\paragraph{Objective.}
The aim of our paper is not to reproduce all the exposed results and explored paths.
Instead, we intend to check some of these results on latest hardware components, and to consider new potential controllable factors that could also contribute to the energy variation.
We also aim to provide practitioners with guidelines to conduct robust experiments in a controlled environment, thus reducing the variation overhead to share trustable results and build credible comparisons.

@@

>>>>

% \section{Introduction}
%\lettrine[nindent=0em,lines=3]{C} 
% To conduct robust evaluations, practitioners often try to ensure reproducible environmental conditions in order to properly benchmark their software systems.
% 
% Recently, the research community has been investigating typical "crimes" in systems benchmarking and established guidelines for conducting robust and reproducible evaluations~\cite{DBLP:journals/corr/abs-1801-02381}.

% In theory, using identical CPU, same memory configuration, similar storage and networking capabilities, should favour reproducible experiments.
% However, when it comes to measuring the energy consumption of a system, applying acknowledged guidelines and carefully repeating the same benchmark can nonetheless lead to different energy footprints not only on homogeneous nodes, but even within a single node.
% This difference---also called \emph{energy variation} (EV)---has become a serious threat to the accuracy of experimental evaluations.

% Figure~\ref{fig:motivation} illustrates this variation problem as a violin plot of $20$ executions of the benchmark \emph{Conjugate Gradient} (\textsf{CG}) taken from the \emph{NAS Parallel Benchmarks} (NBP) suite~\cite{Bailey:1991:NPB:125826.125925}, on $4$ nodes of an homogeneous cluster (the cluster \textsf{Dahu} described in Table~\ref{table:g5k}) at 50\,\% workload.
% One can observe a large variation of the energy consumption, not only among homogeneous nodes, but also at the scale of a single node, reaching up to $25\,\%$ in this example.

% \begin{figure}%[!htb]
%     \center{\includegraphics[width=.9\linewidth]{imgs/motivation}}
%     \caption{CPU energy variation for the benchmark \textsf{CG}}\label{fig:motivation}
% \end{figure}

% TODO Should I add the hardware mitigation  for energy variation ?

@@

Unfortunately, hardware factors can be hardly tuned to tame the energy variation that can be observed in systems benchmarking.
For example, managing the CPU temperature or a server position in a cluster, are not actions that one can easily do, especially on the modern data centers and cloud platforms.
Therefore, the goal of this paper is to investigate the spectrum of factors that can cause or increase the variability of energy consumption in systems benchmarking, and to propose effective guidelines to control such factors in order to mitigate this variability.
% While this paper does not question the benefits of established CPU features, like C-states or Turbo~Boost, it delivers a deeper analysis of the effects they may introduce on a wide set of experiments and nodes.
% By quantifying potential energy variations induced by controllable factors, we intend to identify the proper configurations that minimize energy variations, depending on the workload characteristics.
% These guidelines aim at supporting practitioners in conducting robust system benchmarks and reporting reproducible energy consumption, rather than recommending production-scale practices or solutions to reduce the power consumption of a given system.
% 
% The key contributions of this paper can therefore be summarized as:
% \begin{enumerate}
%     \item Providing a better understanding of the energy variation, by using different generations of CPU deployed in $4$ clusters with more than $100$ physical nodes, and by considering existing systems benchmarks with diverse workloads;
%     \item Identifying controllable factors that contribute to the variation in CPU energy consumption, comparing them against the state of the art, and completing them with other uncovered assumptions;
%     \item Reporting on some guidelines on how to conduct reproducible experiments with less energy variations;
%     \item Discussing the differences between inter-nodes and intra-nodes variations.
% \end{enumerate}

@@

Beyond hardware components, the accuracy of power meters has also been questioned.
Inadomi~\emph{et~al.}~\cite{inadomi_analyzing_2015} used three different power measurement tools: RAPL, Power Insight\footnote{\url{https://www.itssolution.com/products/trellis-power-insight-application}} and BGQ EMON.
All of the three tools recorded the same $10\,\%$ of energy variation, that was supposedly related to the manufacturing process.
%% this is to prove that the problem is REAL !!! 
\paragraph{Mitigating Energy Variations.}
Acknowledging the energy variation problem on processors, some papers proposed contributions to reduce and mitigate this variation.
In~\cite{inadomi_analyzing_2015}, the authors introduced a variation-aware algorithm that improves application performance under a power constraint by determining module-level (individual processor and associated DRAM) power allocation, with up to $5.4\times$ speedup.
The authors of~\cite{hammouda_noise-tolerant_2015} proposed parallel algorithms that tolerate the variability and the non-uniformity by decoupling per process communication over the available CPU.
Acun~\emph{et~al.}~\cite{acun_variation_2016} found out a way to reduce the energy variation on Ivy~Bridge and Sandy~Bridge processors, by disabling the Turbo~Boost feature to stabilize the execution time over a set of processors.
They also proposed some guidelines to reduce this variation by replacing the old slow chips, by load balancing the workload on the CPU cores and leaving one core idle.
They claimed that the variation between the processor cores is insignificant.
In~\cite{chasapis_runtime-guided_2016}, the researchers showed how a parallel system can be used to deal with the energy variation by compensating the uneven effects of power capping.

@@

% \paragraph{Objective.}
% The aim of our paper is not to reproduce all the exposed results and explored paths.
% Instead, we intend to check some of these results on latest hardware components, and to consider new potential controllable factors that could also contribute to the energy variation.
% We also aim to provide practitioners with guidelines to conduct robust experiments in a controlled environment, thus reducing the variation overhead to share trustable results and build credible comparisons.

@@

\subsection{Testing Protocol}
the results of this protocil can be found in the project of the continious integrtation that was made by the intern Mamadou to give a feedback about the evolution of the energyconsumption within time

### 2022-09-03 chapters/benchmarking/benchmarking.tex
refactoring
refactoring : continuation
refactoring : stateoftheart

\newpage

\chapter{Empirical Evaluation Protocol}
\label{chapter:testing}

To optimize the software energy consumption, a reproducible setup environment requires to be designed.
I chose to adopt an empirical approach to explore energy consumption reductions through a trial and error methodology.

%  This chapter will be dedicated to provide a set of guidelines to provide a \em{reproducible, accurate} and \em{representative} tests. Each aspect will be detailed in the sections below.

\import{chapters/testing/}{challenges}
\import{chapters/testing}{stateoftheart}
\import{chapters/testing/}{environement}
\import{chapters/testing/}{experement}
\import{chapters/testing/}{conclusion}

>>>>

\newpage
\chapter{A Benchmarking Protocol to Measure Software Energy Consumption}\label{chapter:benchmarking}

\import{\currfiledir}{draft}
\import{\currfiledir}{reproducibility}
\import{\currfiledir}{experiment}
\import{\currfiledir}{conclusion}

% However for them to be more representatives we wanted to simulate the production enviromenet.
%% should i add  the thing about green faas ? 
%% same about what we have done with python workers and lazy ? basically just a graph

### 2022-09-03 chapters/benchmarking/experiment.tex

\newpage
% To satisfy those three requirements. we elaborated a benchmarking protocol that aims to help reaserchers and practicioners in their experementations.
% for each aspect we present a set of guidelines and tools.
%% This is a conclusion maaan  !! 
\section{Improving accuracy by taming the energy variation}\label{sec:taming-the-energy-variation}

@@

To meet these criteria, we looked at the "state of the art" and found the most common benchmarks used to test the performance of hardware. We then chose the ones that could be scaled up. 
Our candidate is  emph{NAS Parallel Benchmark} (NPB v3.3.1)~\cite{Bailey:1991:NPB:125826.125925}: one of the most used benchmarks for \emph{HPC}. We used the applications (\textsf{LU}), the \emph{Conjugate Gradient} (\textsf{CG}) and \emph{Embarrassingly Parallel} (\textsf{EP}) computation-intensive benchmarks in our experiments, with the \textsf{C} data class.
Furthermore we have used other applications to validate our results using more applications such as \texttt{Stress-ng v0.10.0},\footnote{\url{https://kernel.ubuntu.com/~cking/stress-ng}} \texttt{pbzip2 v1.1.9},\footnote{\url{https://launchpad.net/pbzip2/}} \texttt{linpack}\footnote{\url{http://www.netlib.org/linpack}} and \texttt{sha256 v8.26}\footnote{\url{https://linux.die.net/man/1/sha256sum}}.

>>>>

\newpage
% To satisfy those three requirements. we elaborated a benchmarking protocol that aims to help reaserchers and practicioners in their experementations.
% for each aspect we present a set of guidelines and tools.
%% This is a conclusion maaan  !! 
\section{Improving Accuracy by Taming Energy Variations}\label{sec:taming-the-energy-variation}

@@

To meet these criteria, we looked at the "state of the art" and found the most common benchmarks used to test the performance of hardware. We then chose the ones that could be scaled up. 
Our candidate is  emph{NAS Parallel Benchmark} (NPB v3.3.1)~\cite{Bailey:1991:NPB:125826.125925}: one of the most used benchmarks for \emph{HPC}. We used the applications (\textsf{LU}), the \emph{Conjugate Gradient} (\textsf{CG}) and \emph{Embarrassingly Parallel} (\textsf{EP}) computation-intensive benchmarks in our experiments, with the \textsf{C} data class.
Furthermore we have used other applications to validate our results using more applications such as \texttt{Stress-ng v0.10.0},\footnote{\url{https://kernel.ubuntu.com/~cking/stress-ng}} \texttt{pbzip2 v1.1.9},\footnote{\url{https://launchpad.net/pbzip2/}} \texttt{linpack}\footnote{\url{http://www.netlib.org/linpack}} and \texttt{sha256 v8.26}.\footnote{\url{https://linux.die.net/man/1/sha256sum}}

### 2022-09-03 chapters/literature/draft_citations.tex
refactoring : revert the changes for the benchmarking protocol chapter

Efficiency in energy usage is a well-known topic. In the majority of fields, the purpose is to minimize the energy consumption of electrical devices/components. Modern times even see energy classification (A, B,... F) for Homes, cars and electronic products, to provide the consumer an indication of the energy consumption of his devices, which will reflect on his power bill. This creterions is extended even to the hardware components of computer.
figure\ref{fig:soa_comparaisoncpu} \footnote{\url{https://www.cpubenchmark.net/compare/Intel-i9-12900KS-vs-Intel-i9-12900KF/4813vs4611}}

@@

% We first start by defining the word \textbf{benchmark},  
% definition :benchmark: [noun] something that serves as a standard by which others may be measured or judged.
% that serves as a basis for evaluation or comparison (as of computer system performance)
\subsubsection{Impact of these challenges in the empirical research}
In their work~\cite{van_der_kouwe_benchmarking_2018}, Van-der-Kouwe~\emph{et~al.} investigated 50 papers published in top venues
to find out that Tier-1 papers commit an average of five benchmarking crimes.
To analyze the magnitude of the phenomenon, they have identified a set of 22 "benchmarking crimes" that threaten the validity of the system.

@@

%%%% TODO : Elaborate this section and check what should be added 
% As Stephen M. Blackburn~\emph{et~al.} cited in their paper "evaluate collaboratory"~\cite{stephen_evaluate_2012}, one of the major pitfalls of the measurement contexts is the inconsistency, which can be translated here by the fact that the production context is not the same as the benchmarking one.

@@

PowerMon and its upgraded version powerMon2 \cite{bedard2010powermon} are based on an micro-controller chip that can monitor up to 6 channel (8 for powermon2), simultanetly. 
Therefore, we are able to monitor the power consumption of 4 devices at the same time. 
The frequency sample of this tool is up to 50\,Hz, with an accuracy of 1.2\%.
Powermore2 comes with smaller size that can fit within 3.5 inches rack drive.

>>>>

Efficiency in energy usage is a well-known topic. In the majority of fields, the purpose is to minimize the energy consumption of electrical devices/components. Modern times even see energy classification (A, B,... F) for Homes, cars and electronic products, to provide the consumer an indication of the energy consumption of his devices, which will reflect on his power bill. This criterion is extended even to the hardware components of computer.
Figure~\ref{fig:soa_comparaisoncpu}\footnote{\url{https://www.cpubenchmark.net/compare/Intel-i9-12900KS-vs-Intel-i9-12900KF/4813vs4611}}

@@

% We first start by defining the word \textbf{benchmark},  
% definition :benchmark: [noun] something that serves as a standard by which others may be measured or judged.
% that serves as a basis for evaluation or comparison (as of computer system performance)
\subsubsection{Impact of these Challenges in the Empirical Research}
In their work~\cite{van_der_kouwe_benchmarking_2018}, Van-der-Kouwe~\emph{et~al.} investigated 50 papers published in top venues
to find out that Tier-1 papers commit an average of five benchmarking crimes.
To analyze the magnitude of the phenomenon, they have identified a set of 22 "benchmarking crimes" that threaten the validity of the system.

@@

% %%%%% move it to SOA maybe 
% \subsection{Virtual Machines}
% First option should be using \emph{Virtual Machines} (VM), which gives researchers the freedom to choose the most appropriate tools, software and operating system that they are the most comfortable with, without paying the price to change the actual working environment, which will give them eventually more control over the dependencies and the execution environment.
% Moreover, using a VM addresses the \emph{replication crisis} thanks to the virtual images, as even the most complex architecture can be reproduced easily by instantiating a copy of the image.
% Since the virtual machines are agnostic to the host architecture, researchers will not have to worry about where and how their experiments are replicated, because they have already setup the execution environment.
% Another advantage of VM is the snapshot mechanism, as it allows researchers to create backups and revert some changes with simple clicks. 
% Last but not least, thanks to the isolation, virtual machines push the reproducibility further by allowing the future usages to see all the variables---controlled and uncontrolled---and do other analysis without dealing with any dependencies.
% In his paper~\cite{howe_virtual_2012}, Bill Howe lists the benefits of using VM in researchers experiments, including the economical impact and cultural limitation to a such approach.
% In particular, VM allow them to have control over the resources, the dependencies, and the execution environment.
% Moreover, thanks to snapshots, deploying a software is easily achieved by instantiating a copy of that image.
% However, this choice comes with a certain cost.
% Because of the hypervisor, software will build on two kernels---the virtual machine (guest) one and the host machine one---which will provide a noticeable overhead, and will impact the performances of the system-under-test.
% Therefore, we cannot use VM for experiments that are related to performance.
% Another limitation of VM is the the isolation.
% While this feature prevents the experimental environment from any undesirable interference from the outside world, this interaction may be required---especially when the experiment is dependent to an external 
%%%% TODO : Elaborate this section and check what should be added 
% As Stephen M. Blackburn~\emph{et~al.} cited in their paper "evaluate collaboratory"~\cite{stephen_evaluate_2012}, one of the major pitfalls of the measurement contexts is the inconsistency, which can be translated here by the fact that the production context is not the same as the benchmarking one.

@@

PowerMon and its upgraded version powerMon2 \cite{bedard2010powermon} are based on an micro-controller chip that can monitor up to 6 channel (8 for powermon2), simultaneously. 
Therefore, we are able to monitor the power consumption of 4 devices at the same time. 
The frequency sample of this tool is up to 50\,Hz, with an accuracy of 1.2\%.
Powermore2 comes with smaller size that can fit within 3.5 inches rack drive.

### 2022-09-03 chapters/optimization/java-dir/java.tex

% \section*{JAVA}
\section{Introduction}
As reported in the state of the art, Java is one of the most popular programming languages adopted by practitioners. %ROMAIN: Add source
Furthermore, if we take into consideration legacy applications, Java becomes the most used programming language. %ROMAIN:  Add source
In addition to its popularity, Java exhibits an interesting behavior when it comes to energy consumption and performance, Java applications can be at the same time one of the most energy-efficient or hungry solutions.
As we have seen in the previous chapter, an inappropriate combination of parameters can drive Java applications from the top language to the bottom just by setting the wrong parameters.
in this chapter, we want to dig deeper into this aspect of Java and study its runtime.
This chapter thus focuses on the impact of the runtime of Java applications on energy consumption.

@@

To answer those research questions, we conduct an empirical study to highlight the impact of this runtime.

% \subsection{definition of the JVM}
% TODO : ADD definition here 
\begin{figure}%[ht]
    \centering
    \includegraphics[width=\linewidth]{imgs/Jvm_architecture}
    \caption{JVM architecture}
    \label{fig:hotspot}
\end{figure}

>>>>

% \section*{JAVA}
\section{Introduction}
As reported in the state of the art, Java is one of the most popular programming languages adopted by practitioners.In fact, in 2022, Java will be second only to Python, according to PYPY~\footnote{\url{https://pypl.github.io/PYPL.html}}.
Furthermore, if we take into consideration legacy applications, Java becomes the most used programming language.According to the TIOBE index, Java was the most frequently used language from 2002 until 2017, and it remained in the top 5 after that~\footnote{\url{https://www.tiobe.com/tiobe-index/}}.
In addition to its popularity, Java exhibits an interesting behavior when it comes to energy consumption and performance, Java applications can be at the same time one of the most energy-efficient or hungry solutions.
As we have seen in the previous chapter, an inappropriate combination of parameters can drive Java applications from the top language to the bottom just by setting the wrong parameters.
in this chapter, we want to dig deeper into this aspect of Java and study its runtime.
This chapter thus focuses on the impact of the runtime of Java applications on energy consumption.

@@

To answer those research questions, we conduct an empirical study to highlight the impact of this runtime.
Figure

% \subsection{definition of the JVM}
% TODO : ADD it to the introduction to 
% \begin{figure}%[ht]
%     \centering
%     \includegraphics[width=\linewidth]{imgs/Jvm_architecture}
%     \caption{JVM architecture}
%     \label{fig:hotspot}
% \end{figure}

### 2022-09-03 chapters/optimization/python-dir/python.tex

\section{Green Web Development}\label{sec:webdev}
% TODO : ADD introduction to the orms 
We start by studying the impact of the \emph{Object Relational Mapping} (ORM)~\cite{o2008object} on energy consumption.
% We have chosen Django as our web framework.
The idea behind this use case is to observe if the choice of the database and the ORM impacts the energy consumption and the performance of the website.
We, therefore, create a Django website to analyze the cost of a single request using various mechanisms to retrieve data from the database.

We considered using two different databases, \textsc{PostgreSQL} and \textsc{SQlite3}, that store the same records, and three different ways to fetch the records.
\begin{enumerate}
    \item \textsf{Vanilla} relies only on the ORM to retrieve the data,
    \item \textsf{Prefetch} queries the data before being requested,
    \item \textsf{Optimized} leverages SQL without passing by the ORM.
\end{enumerate}

@@

@@

%%% putting the idea, later i ll reformulate it 
The reason behind such a behavior is mainly related to how the Python interpreter is implemented.
% add refenece https://www.python.org/doc/essays/list2str/ 
%% https://www.pythonpool.com/for-vs-while-loop-python/  
% https://wiki.python.org/moin/TimeComplexity
%% http://python-history.blogspot.com/2010/06/from-list-comprehensions-to-generator.html?m=1
To reduce the latency of Python applications, most of the built-in functions and operations are written in C, and the same goes for the function \texttt{range}.
Furthermore, the function \texttt{len} has a complexity of $\mathcal{O}(1)$ as it is based on the function \texttt{Py\_SIZE} of C, which stores the length in a field for the object.
Therefore, the \texttt{for in range} is creating a new iterator that has the same length as the first one and, for each iteration, requires a second access (\texttt{l[i]}) instead of one---explaining the doubled time.
The \texttt{while} is even slower due to the implicit increment of the variable, which causes an extra operation during the loop.
To confirm this hypothesis, we tried to construct a new list by editing the elements of the previous one (cf. \Cref{fig:pythontreatement}).
And, as predicted, the built-in methods are the most energy-saving ones, while the customize \texttt{while} loop is the heaviest.

>>>>

\section{Green Web Development}\label{sec:webdev}
Django~\footnote{\url{https://www.djangoproject.com/}} and Flask\footnote{\url{https://flask.palletsprojects.com/}}  are the most popular frameworks for web development. According to a Jetbrains poll\footnote{\url{https://lp.jetbrains.com/python-developers-survey-2021/}} , Django is used in 40\% of the cases, while Flask is used in 41\% of the cases.

In contrast to Flask, which is still a microframework, Django is a high-level web framework that provides a standard method for creating and maintaining complex and scalable database-driven websites quickly and effectively. Therefore, this study will focus on The latter one.

\subsection{life cycle of a request in Django}

Django is a MVT (Model-View-Template) framework, which means that it follows the MVC (Model-View-Controller) pattern. Figure~\ref{fig:django-life-cycle} describes the life cycle of a request in this framework.  Whenever a request arrives in Django, it is processed by several middlewares, one at a time. These middlewares are in charge of authentication, security, and so on.
Once the request is processed by these middlewares, it is passed to the URL Router, which will simply extract the URL from the request and will try to match it to the defined URLs.
After getting the matching URL, the corresponding view function is called. This function is responsible for treating the request, gathering the data, and then generating the response that will be put inside a template to be returned as a response.
However, as mentioned before, Django is an MVT model. which offers an automatic way to retrieve data from the database to the view using the \emph{Object Relational Mapping} (ORM)~\cite{o2008object}.

\begin{figure}[hbt]
    \centering
    \includegraphics[width=\linewidth]{imgs/django_request_lifecycle}
    \caption{Request-Response life cycle in Django}
    \label{fig:django-life-cycle}
\end{figure}

First, we start by investigating the energy consumption of the request-response life cycle in Django, in ordrer to determine which part consumes the most energy. To do so, we created a simple Django application that simply returns the response of the request. and then tracked its energy consumption using joulehunter~\footnote{\url{https://github.com/powerapi-ng/joulehunter}}.

Joulehunter is an open source library that we developed in order to help practitioners find the energy hotspots in their applications using statistical profilers. In the case of Django, it can be added as middleware without any extra setup or changes to the source code. The energy consumption of the request-response life cycle in Django is shown in the figure~\ref{fig:django_life_cycle_naive}.As we can notice, 91.4\% of the total energy consumption is spent to resolve the request by retrieving the data while only 5.27\% is spent to render the response.

\begin{figure}[hbt]
    \centering
    \includegraphics[width=\linewidth]{imgs/django_life_cycle_naive}
    \caption{Tree representation of the energy consumption of a single request in django (naive version)}
    \label{fig:django_life_cycle_naive}
\end{figure}

%% REMARQUE the experiment with joulehunter isn't conducted within the same machine that is used for generating the boxplot graphs 
% We have chosen Django as our web framework.
Therefore,we chose to observe if the choice of the database and the ORM impacts the energy consumption and the performance of the website.
To accomplish so, we examined the cost of a single request of the prior website using various ways to extract data from the database.
We considered using two different databases, \textsc{PostgreSQL} and \textsc{SQlite3}, that store the same records, and three different ways to fetch the records.
\begin{enumerate}
    \item \textsf{Vanilla} relies only on the ORM to retrieve the data,
    \item \textsf{Prefetch} queries the data before being requested,
    \item \textsf{Optimized} leverages SQL without passing by the ORM.
\end{enumerate}

@@

Finally, we run the same experiment with the \textsf{Optimized} strategy 
using joulehunter
.Figure~\ref{fig:django_profiled_optimized} shows the new graph of the energy consumption. As one can notice. while the rendering method consumed the same amount of energy as in  the \textsf{Vanilla} strategy (arround 1.3KJ), the resolve part droped by $20\times$.

\paragraph{Conclusion}
We conclude this study by stating that the database and ORM selection have a major impact on the website's performance and energy consumption.
And unlike the rendering portion, there are many options available for the database, the ORM, and the data handling strategy.
\begin{figure}[hbt]
    \centering
    \includegraphics[width=\linewidth]{imgs/django_profiled_optimized}
    \caption{Tree representation of the energy consumption of a single request in django (naive version)}
    \label{fig:django_profiled_optimized}
\end{figure}

@@

%%% putting the idea, later i ll reformulate it 
The reason behind such a behavior is mainly related to how the Python interpreter is implemented\footnote{\url{https://www.python.org/doc/essays/list2str}},\footnote{\url{https://www.pythonpool.com/for-vs-while-loop-python/}}.  
To reduce the latency of Python applications, most of the built-in functions and operations are written in C, and the same goes for the function \texttt{range}.\footnote{\url{ http://python-history.blogspot.com/2010/06/from-list-comprehensions-to-generator.html?m=1}}
Furthermore, the function \texttt{len} has a complexity of $\mathcal{O}(1)$ as it is based on the function \texttt{Py\_SIZE} of C, which stores the length in a field for the object~\footnote{\url{https://wiki.python.org/moin/TimeComplexity}}.
Therefore, the \texttt{for in range} is creating a new iterator that has the same length as the first one and, for each iteration, requires a second access (\texttt{l[i]}) instead of one---explaining the doubled time.
The \texttt{while} is even slower due to the implicit increment of the variable, which causes an extra operation during the loop.
To confirm this hypothesis, we tried to construct a new list by editing the elements of the previous one (cf. \Cref{fig:pythontreatement}).
And, as predicted, the built-in methods are the most energy-saving ones, while the customize \texttt{while} loop is the heaviest.

### 2022-09-03 chapters/programming_languages/programming_languages.tex

\chapter{Energy footprint of programming languages}
\label{chapter:porgramming_langauges}
In this chapter, we talk about how the programming language affects how much energy the software consumes.We suggest starting with general microbenchmarking and watching how each programming language performs with the CPU and memory.The main goal of this chapter is to advise developers on how to choose a programming language based on their project's needs in order to make their product use the least amount of energy possible. For such a question, no answer is evidedent.But there are some features we can take from each programming language, such as: 
\begin{itemize}
    \item performance
    \item community support
    \item scalability
    \item energy consumptiom
    \item memory usage
\end{itemize}

As we saw in the last chapter, one of the most important things about a test is how well it \textsc{represents} the production environment. Therefore, we extend this study to include real-world use cases, with two case studies provided in the parts that follow. 
\section{Remote Procedure Call Frameworks}
\import{\currfiledir/rpc}{rpc}
\section{web Frameworks}
\import{\currfiledir/web_frameworks}{web_frameworks}

>>>>

\chapter{Energy Footprint of Programming Languages}
\label{chapter:porgramming_langauges}
In this chapter, we talk about how the programming language affects how much energy the software consumes.We suggest starting with general micro-benchmarking and watching how each programming language performs with the CPU and memory.The main goal of this chapter is to advise developers on how to choose a programming language based on their project's needs in order to make their product use the least amount of energy possible. For such a question, no answer is obvious.
Nonetheless, there are some features we can take from each programming language, such as: 
\begin{itemize}
    \item performance,
    \item community support,
    \item scalability,
    \item energy consumption,
    \item memory usage.
\end{itemize}

As we saw in the last chapter, one of the most important things about a test is how well it \textsc{represents} the production environment.
Therefore, we extend this study to include real-world use cases, with two case studies provided in the parts that follow.

\section{Investigating Remote Procedure Call Frameworks}
\import{\currfiledir/rpc}{rpc}

\section{Investigating Web Application Frameworks}
\import{\currfiledir/web_frameworks}{web_frameworks}
